Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars

### 2014-10-21 drafts/cfp/draft.tex
Finished intro draft.

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{Shannon1950}, an adult English speaker's understanding of
her language and her world is such that the entropy of a character in
English text is slightly more than one bit. Plausibly, any machine
learning system able to predict natural language text with such
accuracy will have learned much about the concepts expressed in that
text. This intuition motivates us to invert Shannon's experiment and
learn linguistic concepts by learning a probabilistic generative model
for natural language.

Probabilistic Context Free Grammars (PCFGs) are popular and much
studied generative models for natural language
syntax~\cite{Someone}. While debate continues about their ability to
express the syntax of natural language, it is agreed that they cannot
express dependencies between words that depend upon each other for
their meaning. Consider an example from the domain of sentences
addressed in this paper: ``after twenty eight comes twenty nine.''
The second ``twenty'' depends on the first ``twenty'' and the
``eight'', and the ``nine'' depends on the ``eight''. This pattern is
a clear example of the \emph{cross-serial dependencies} which
context-free grammars cannot support.

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\cite{SOMEONE} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

To explore this approach, we introduce a probabilisitic model called a
Latent Predicate Network (LPN), and describe an inference procedure to
learn its parameters using the probabilistic logic programming
language PRISM. Finally, we study its performance when trained on
sentences describing the relationships between numbers.

>>>>

Although concept learning and language acquisition have typically been
treated as distinct problems in AI, linguistics and cognitive
development, for a child learning to understand language, they are
strongly coupled. People learn about many abstract concepts primarily
through language and understanding language depends on understanding
concepts.  Research in concept learning is often focused on concepts
grounded in perceptual features, and while it is probably true that
many concepts are learned via generalization from examples, some
concepts cannot possibly be learned this way.

Number concepts are a good example: children do not learn about the
meaning of ``seventy five'' by seeing examples of seventy five things;
they do not know that ``seventy five'' is more than ``twenty five''
because of their perceptual experiences of these quantities. Rather,
children learn the meaning of ``seventy five'' (or ``a billion and
five'') by noticing how number words are used in language, in counting
sequences, in arithmetic excercises, etc. Other good examples of such
abstract concepts are kinship and social relations (e.g. ``my father
in law's grandmother''), temporal relations (``the day after last
Thanksgiving.''), and spatial relations (``above and just to the left
of'').

Such abstract concepts share many of the properties of language
syntax: there is an unbounded set of them, they derive their
meanings via composition, and, although people only ever say, hear, or
read about a small number of them, they are able to reason correctly
about any them. There seems to be a grammar to these concepts, and
grasping this grammar is critical to understanding their meanings and
how to use them. This motivates our approach here, which is to apply
the tools of probabilistic grammars more familiar from studies of
syntax to the problem of concept aquisition.

Doing so requires overcoming some technical barriers.

First, whereas context-free grammars are suitable for describing large
swaths of language syntax, the grammars of concepts are not
context-free. To address this, we use Range Concatentation Grammars, a
context-sensitive grammar formalism -- one of several developed within
linguistics -- and extend this formalism to a probabilistic model of
sentences.

Second, the categories of syntax -- the nonterminals of the grammar --
are often assumed to be known to children innately and given to
automated learners by human experts. The categories that underlie
conceptual knowledge, on the other hand, are far more numerous, vary
from domain to domain, and are unlikely to be known to the
learner. This motivates our use of latent predicates that, through
learning, assume the role of a domain's underlying concepts (in the
number domain, these might correspond to the concepts of
successorship, order of magnitude, magnitude comparison,
exact vs. approximate, etc.).

Finally, using probabilistic context-sensitive grammars with latent
predicates brings up problems of tractability: our goal is to find a
middle ground between expressivity and tractability. Using tools like
PRISM -- a probabilistic logic programming system that naturally
implements efficient dynamic programming algorithms for our models --
we are able to explore which domains and which grammar architectures
are a good fit for this grammar-based approach.

In the rest of this paper describe early work on this approach. First, we
present our probabilistic model of concept learning, which we call
Latent Predicate Networks (LPNs). Then  we describe our approach to
inference and its implementation in PRISM. Finally, we present
preliminary experimental results in the domain of number concept
learning that demonstrate the promise of this approach.

% people learn about many
% abstract concepts primarily through language and, conversely, these
% abstract concepts are critical to understanding language. Consider a
% child learning the meaning of the number words: the meaning of the
% ``seventy five'' is learned not by seeing collections of seventy five
% things but rather by understanding how ``seventy five'' is used in the
% various sentences, counting sequences, arithmetic excercises, etc.

% As Claude Shannon demonstrated with his
% guessing game experiment in 1950~\cite{Shannon1950}, an adult English
% speaker's understanding of her language and her world is such that the
% entropy of a character in English text is slightly more than one
% bit. Plausibly, any machine learning system able to predict natural
% language text with such accuracy will have learned much about the
% concepts expressed in that text. This intuition motivates us to invert
% Shannon's experiment and learn linguistic concepts by learning a
% probabilistic generative model for natural language.

% Probabilistic Context Free Grammars (PCFGs) are popular and much
% studied generative models for natural language
% syntax~\cite{Someone}. While debate continues about their ability to
% express the syntax of natural language, it is agreed that they cannot
% express dependencies between words that depend upon each other for
% their meaning. Consider an example from the domain of sentences
% addressed in this paper: ``after twenty eight comes twenty nine.''
% The second ``twenty'' depends on the first ``twenty'' and the
% ``eight'', and the ``nine'' depends on the ``eight''. This pattern is
% a clear example of the \emph{cross-serial dependencies} which
% context-free grammars cannot support.

% Our approach, then, will be to explore the use of probabilistic
% context \emph{sensitive} grammars for modeling natural language,
% extending the tools of hierarchical grammar induction~\cite{SOMEONE} from
% context-free grammars to ones that can express not only syntactic
% relationships but semantic ones as well. In the place of latent
% syntactic categories, we will refer to latent predicates -- relations
% between sentence components -- that, when learned, should correspond
% to the concepts that underlie sentence semantics.

% To explore this approach, we introduce a probabilisitic model called a
% Latent Predicate Network (LPN), and describe an inference procedure to
% learn its parameters using the probabilistic logic programming
% language PRISM. Finally, we study its performance when trained on
% sentences describing the relationships between numbers.
